---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Student Name"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)
library(here)
here()

chem_physics <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

chem_physics$sampledate <- as.Date(chem_physics$sampledate, format = "%m/%d/%y")

str(chem_physics$sampledate)
head(chem_physics$sampledate)

#2
my_theme <- theme_minimal() +
  theme(
    # text
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
    axis.title.y = element_text(size = 12, margin = margin(r = 8)),
    axis.text = element_text(size = 10),
    # lines & grids
    axis.ticks = element_line(linewidth = 0.3),
    panel.grid.major = element_line(linewidth = 0.2, linetype = "dashed", color = "grey75"),
    panel.grid.minor = element_blank(),
    # backgrounds
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = "grey85"),
    legend.key = element_rect(fill = "white", color = NA),
)

# Set as default theme
theme_set(my_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July did not change with depth across all lakes. 
Ha: The mean lake temperature recorded during July did change with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
chem_physics_july <- chem_physics %>%
  filter(format(sampledate, "%m") == "07") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#5
# Scatterplot
temptbydepth <-
  ggplot(chem_physics_july, aes(x = depth, y = temperature_C)) +
  ylim(0, 35) +
  geom_point()+
  my_theme
print(temptbydepth)

# Add a line and standard error for the linear regression
temptbydepth1 <- 
  ggplot(chem_physics_july, aes(x = depth, y = temperature_C)) +
  geom_smooth(method = "lm") +
  ylim(0, 35) +
  geom_point() 
print(temptbydepth1) 

# SE - confidence interval around smooth can also be removed
temptbydepth2 <- 
  ggplot(chem_physics_july, aes(x = depth, y = temperature_C)) +
  ylim(0, 35) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = "black")
print(temptbydepth2)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Temperature decreases steadily with depth, showing a clear negative trend. The points mostly follow a straight line, though some curvature suggests natural layering in the lake.



7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
# Linear regression test
tempdepth_lm <- lm(data = chem_physics_july, temperature_C ~ depth)

# Display results
summary(tempdepth_lm)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model shows a significant relationship between temperature and depth. About 73.87% of the variation in temperature is explained by changes in depth (R^2 ~= 0.7387). Based on 9,726 degrees of freedom, this relationship is highly significant (p < 2.2e-16). Temperature is predicted to drop by about 1.95 °C for every 1 m increase in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
tempdepth_AIC <- lm(data = chem_physics_july, temperature_C ~ year4 + daynum + depth)

#Choose a model by AIC in a Stepwise Algorithm
step(tempdepth_AIC) # none = 26066: lowest AIC when dropping none

#10
tempdepth_AIC <- lm(data = chem_physics_july, temperature_C ~ year4 + daynum + depth)
summary(tempdepth_AIC)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final model includes year4, daynum, and depth as predictors of temperature. Together they explain about 74.11% of the variation in temperature, showing a strong fit (p < 2.2e-16) . This is only a slight improvement over the model with depth alone, suggesting depth is the main factor, with minor effects from year and day.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Format ANOVA as aov
temp_anova <- aov(temperature_C ~ lakename, data = chem_physics_july)
summary(temp_anova)

# Format ANOVA as lm
temp_lm <- lm(temperature_C ~ lakename, data = chem_physics_july)
anova(temp_lm)
summary(temp_lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The ANOVA shows a significant difference in mean July temperature among lakes (F(8, 9719) = 50.0, p < 2.2e-16). The linear model indicates that several lakes, including East Long, Hummingbird, Tuesday, and West Long, are cooler than the reference lake, while others like Crampton and Paul are slightly cooler. Lake ID explains about 3.8% of the variation in temperature, which suggest that local lake characteristics influence water temperatures in July.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

lake_temp_plot <- 
  ggplot(chem_physics_july, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, size = 1.2) +
  geom_smooth(aes(group = lakename), method = "lm", se = FALSE, linewidth = 0.9) +
  ylim (0, 35) +                    
  labs(x = "Depth (m)", y = "Temperature (°C)",
       title = "July Temperature–Depth by Lake") +
  my_theme
lake_temp_plot

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# Post-hoc test
# TukeyHSD() computes Tukey Honest Significant Differences
TukeyHSD(temp_anova)

# Extract groupings for pairwise relationships
temp_groups <- HSD.test(temp_anova, "lakename", group = TRUE)
temp_groups

# Graph the results

tempbylake_boxplot <- ggplot(temp_anova, aes(x = lakename, y = temperature_C)) +
  geom_boxplot() +
  my_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("a", "ab", "e", "de", "c", "c", 
                         "de", "bc", "d")) +
  labs(x = "Lake", y = "Temperature (°C)") +
  ylim(0, 35)
tempbylake_boxplot
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Based on the Tukey HSD results, Peter Lake (group “c”) shares statistically similar mean temperatures with Paul Lake and Ward Lake, as they belong to overlapping groups that include the letter “c.” This means there is no significant difference in temperature among these three lakes. There is no lake that is statistically distinct from all the other lakes. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can use two-sample t-test to directly test whether their mean temperatures differ significantly.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Filter data for Crampton Lake and Ward Lake
cramptonvsward <- chem_physics_july %>%
  filter(lakename == "Crampton Lake"|lakename =="Ward Lake")

#Format as a t-test
cramp_ward_t.test <- t.test(temperature_C ~ lakename, data = cramptonvsward)
cramp_ward_t.test
```

>Answer: The t-test (p = 0.26) shows no significant difference in July mean temperature between Crampton Lake (15.35°C) and Ward Lake (14.46°C). Their means are statistically similar, matching the earlier Tukey HSD result that grouped them together.

